Measurement Clusterization in Bayesian D-optimal Designs in Infinite Dimensions

### 2024-07-31 latex/vanishing.tex
vanishin almost done

\begin{lemma}[Simultaneous diagonizability]\label{lemma:sim_diag}
  Let $\hil$ separable Hilbert space, $C:\hil \to \hil$ self-adjoint
  and $\func_1,\dots,\func_m \in \hil$. Denote $\func^*$ the element
  $\func$ acting as a linear functional. If
  \begin{equation*}
   (C + \sum_{j=1}^m \func_j\func_j^*) \func_l = \xi_l \func_l,\ l = 1,\dots,m
  \end{equation*}
  then $C$ and $\sum_{j=1}^m \func_j \func_j^*$ are simultaneously
  diagonalizable.
\end{lemma}
The proof of Lemma \ref{lemma:sim_diag} is delegated to the
Supplementary.

@@

The proof of Lemma is also delegated to the Supplementary. It is
however important to note that this proof is construcive. It will give
us a way to construct D-optimal designs, once we fully characterize
them in Theorem \ref{thm:char} below.

@@

  Then:
  \begin{enumerate}
  \item  \(\tr{\opt^*\opt} = m\).
  \item \(\opt^*\opt\) and \(\fwd\prcov\fwd^*\) are simultaneously
    diagonalizable.
  \item \(k := \rank \opt^*\opt \leq m\) and
    \begin{equation*}
      \tar(\opt) = \frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i). %= \frac12 \sum_{i=1}^{m} \log (1 + \sigma^{-2}\lambda_i\eta_i).
    \end{equation*}
  %% \item 
  %%   \begin{equation*}
  %%     k = \argmax \left \{ k:\lambda_k^{-1} < \sigma^{-2}\frac{m}{k} + \frac{1}{k} \sum_{j=1}^{k}
  %%     \lambda_j^{-1} \right \}.
  %%   \end{equation*}
  \item
    \begin{equation*}
        \eta_i = \begin{cases}
          \frac{m}{k} - \sigma^2 \lambda_i^{-1} + \sigma^2 \frac{1}{k} \sum_{j=1}^k \lambda_j^{-1} & 1 \leq i \leq k \\
          0 & i > k 
        \end{cases}.
    \end{equation*}
  \item The covariance of the pushforwad \(\fwd_{*} \postopt\) is \(\left
    ( (\fwd \prcov \fwd^*)^{-1} + \sigma^{-2} \opt^*\opt \right
    )^{-1}\) and its eigenvalues are
    \begin{equation}\label{eq:cylinders}
      \theta_i =
      \begin{cases}
        \left(\frac{\sum_{j=1}^k \lambda_j^{-1} + \sigma^{-2}m}{k} \right )^{-1} & i \leq k \\
        \lambda_i &  i > k 
      \end{cases}
    \end{equation}
  \end{enumerate}
\end{theorem}

@@

\subsection{Answer to Question \ref{q:why}}\label{subsec:why}
Theorem \ref{thm:char} helps us give a compelling explanation of the
measurement clusterization we obeserved for the inverse problem of the
heat equation (Fig.~\ref{fig:}). Below, we take the following steps:
We first utilize Theorem \ref{thm:char} to understand what a D-optimal
design will look like for our relaxed model; we then argue that
although not identical, a D-optimal design for the full inverse
problem of the heat equation shares similar charachteristics to the
relaxed design; finally, we show how ...

First, recall the relevant parts of Theorem \ref{thm:char}. In our
relaxed model, for a D-optimal design $\opt$:
\begin{itemize}
  \item $\opt^*\opt$ and $\fwd\prcov \fwd^*$ are simultaneously
    diagonalizable.
  \item $\opt^*\opt$ has non-zero eigenvalues only for the $k:=\rank
    \opt^*\opt$ largest eigenvalues $\lambda_j$ of $\fwd\prcov\fwd^*$,
    i.e.~$\eta_j=0$ for $j>k$.
  \item $k$ is monotonic in $m$, the number of measurements we take.
\end{itemize}

Consider $\fwd$ and $\prcov$ from the \emph{the inverse problem of the
heat equation}. As before, we denote the eigenvalues of
$\fwd\prcov\fwd^*$ by $\lambda_j$. We input these eigenvalues into our
\emph{relaxed} model, and find a D-optimal design $\opt$ for our
relaxed model, e.g.~by using Lemma \ref{lemma:free} in the proof of
Theorem \ref{thm:char}. In our relaxed model, the measurements we take
are best utilized in reducing uncertainty for the first $k$
eigenvectors. So, a D-optimal design arising from our \emph{relaxed
model} completely avoids measuring eigenvectors $k+1$ and above.

Of course, in a real life problem --- such as the inverse problem of
the heat equation --- it is likely impossible to find measurement
locations for which all eigenvectors $k+1$ and above are
zero. Luckily, the eigenvalues of $\fwd\prcov\fwd^*$ decay very
quickly: for example, recall from Section \ref{subsec:toy} that the
eigenvalues of the forward model for the 1D heat equation are $-(2\pi
n)^{-2}$. Thus a D-optimal design is a trade-off between measuring the
small number of eigenvectors we are interested in.

We explore the abovementione trade-off for the inverse problem of the
heat equation in Fig.~\ref{fig:eigenvectors}. We allow $m=4$
measurements in $\Omega = [0,1]$ and observe that D-optimal
measurement locations are clustered at $x_1 = 0.31$ and $x_2 =
0.69$. Upon close inspection of the scaled eigenvectors of $\fwd
\prcov \fwd^*$, we first observe that eigenvectors $3$ and above have
negligible prior amplitude. Since we only have $m=4$ measurements at
our disposal, we interpret these results, following Theorem
\ref{thm:char}, as implying we should only care about measuring the
first and second eigenvectors. Then, we note the D-optimal $x_1,x_2$
present a compromise between the amplitude of the first and second
eigenvectors. For example, a measurement at $x=0.5$ would have ignored
the second eigenvector altogether, since the second eigenvector is
zero at $x=0.5$.  Now we can understand measurement clusterization:
attempting to measure what we want and ignoring the rest, along with
the spatial limitations on measurements imposed by the structure of
eigenvectors of $\fwd \prcov \fwd^*$ is the cause for measurement
clusterization in D-optimal designs: we have $m=4$ measuremnts at our
disposal but only two spatial locations that are a good compromise
between the scaled eigenvectors. Thus, clusterization arises by the
pigeonhole principle.

\begin{figure}\label{fig:eigenvectors}
    \centering
    \includegraphics[width=\textwidth]{figs/eigenvectors_dst_scaled.png}
    \caption{D-optimal measurement locations ($m=4$ measurements) and
      weighted eigenvectors for finding the initial condition of the
      1D heat equation. Measurement locations and weighted
      eigenvectors are plotted over the computational domain $\Omega =
      [0, 1]$ (x-axis). Measurement clusterization occurs
      approximately at $0.31$ and $0.69$. These two locations are a
      compromise between the magnitudes of the first and second
      eigenvectors. These are the eigenvectors a D-optimal design aims
      to measure. The rest of the eAllocating $m=4$ measurements into two locations
      results in clusterization, according to the pigeonhole
      principle.}
  \label{fig:why}
\end{figure}

>>>>

\begin{lemma}[Simultaneous diagonalizability]\label{lemma:sim_diag}
  Let $\hil$ separable Hilbert space, $C:\hil \to \hil$ self-adjoint
  and $\func_1,\dots,\func_m \in \hil$. Denote $\func^*$ the element
  $\func$ acting as a linear functional. If
  \begin{equation*}
   (C + \sum_{j=1}^m \func_j\func_j^*) \func_l = \xi_l \func_l,\ l = 1,\dots,m
  \end{equation*}
  then $C$ and $\sum_{j=1}^m \func_j \func_j^*$ are simultaneously
  diagonalizable.
\end{lemma}
The proof of Lemma \ref{lemma:sim_diag} is delegated to the
Supplementary.

@@

The proof of Lemma \ref{lemma:free} is also delegated to the
Supplementary. It is however important to note that this proof is
constructive; it will allow us to construct D-optimal designs, once we
fully characterize them in Theorem \ref{thm:char} below.

@@

  Then:
  \begin{enumerate}
  \item  \(\tr{\opt^*\opt} = m\).
  \item \(\opt^*\opt\) and \(\fwd\prcov\fwd^*\) are simultaneously
    diagonalizable.
  \item \(k := \rank \opt^*\opt \leq m\) and
    \begin{equation*}
      \tar(\opt) = \frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i). %= \frac12 \sum_{i=1}^{m} \log (1 + \sigma^{-2}\lambda_i\eta_i).
    \end{equation*}
  %% \item 
  %%   \begin{equation*}
  %%     k = \argmax \left \{ k:\lambda_k^{-1} < \sigma^{-2}\frac{m}{k} + \frac{1}{k} \sum_{j=1}^{k}
  %%     \lambda_j^{-1} \right \}.
  %%   \end{equation*}
  \item
    \begin{equation*}
        \eta_i = \begin{cases}
          \frac{m}{k} - \sigma^2 \lambda_i^{-1} + \sigma^2 \frac{1}{k} \sum_{j=1}^k \lambda_j^{-1} & 1 \leq i \leq k \\
          0 & i > k 
        \end{cases}.
    \end{equation*}
  \item The covariance of the pushforward \(\fwd_{*} \postopt\) is \(\left
    ( (\fwd \prcov \fwd^*)^{-1} + \sigma^{-2} \opt^*\opt \right
    )^{-1}\) and its eigenvalues are
    \begin{equation}\label{eq:cylinders}
      \theta_i =
      \begin{cases}
        \left(\frac{\sum_{j=1}^k \lambda_j^{-1} + \sigma^{-2}m}{k} \right )^{-1} & i \leq k \\
        \lambda_i &  i > k 
      \end{cases}.
    \end{equation}
  \end{enumerate}
\end{theorem}

@@

\subsection{Answer to Question \ref{q:why}}\label{subsec:why}
Building on Theorem \ref{thm:char}, we can now give a compelling
explanation to the measurement clusterization we observed for the
inverse problem of the heat equation (Fig.~\ref{fig:eigenvectors}).

Consider $\fwd$ and $\prcov$ from \emph{the inverse problem of the
heat equation}. As before, we denote the eigenvalues of
$\fwd\prcov\fwd^*$ by $\lambda_j$. We input these eigenvalues into our
\emph{generic} model, and find a D-optimal design $\opt$ for our
generic model using Theorem \ref{thm:char}. In our generic model, the
measurements we take are best utilized in reducing uncertainty for the
first $k$ eigenvectors. So, a D-optimal design arising from our
\emph{generic model} completely avoids measuring eigenvectors $k+1$
and above.

Of course, in a real life problem --- such as the inverse problem of
the 1D heat equation --- it is likely impossible to find measurement
locations for which all eigenvectors $k+1$ and above are
zero. However, if the eigenvalues of $\fwd\prcov\fwd^*$ decay quickly
(recall the square-exponential decay for eigenvalues of the 1D heat
equation in eq.\eqref{eq:decay}), a D-optimal design will try to
balance measuring a small number (i.e.~$k$) of the leading
eigenvectors.

The abovementioned trade-off is explored in
Fig.~\ref{fig:eigenvectors}. We allow $m=4$ measurements in $\Omega =
[0,1]$ and observe that D-optimal measurement locations are clustered
at $x_1 = 0.31$ and $x_2 = 0.69$. Upon close inspection of the scaled
eigenvectors of $\fwd \prcov \fwd^*$, we first observe that
eigenvectors $3$ and above have negligible prior amplitude. Since we
only have $m=4$ measurements at our disposal, we interpret these
results, following Theorem \ref{thm:char}, as implying we should only
care about measuring the first and second eigenvectors. Then, we note
the D-optimal $x_1,x_2$ present a compromise between the amplitude of
the first and second eigenvectors. For example, a measurement at
$x=0.5$ would have ignored the second eigenvector altogether, since
the second eigenvector is zero at $x=0.5$.

Now we can understand measurement clusterization for the inverse
problem of the heat equation. A D-optimal design attempts to measure
the first $k$ eigenvectors of $\fwd \prcov \fwd^*$. But there may be
(spatial) limitations on where these $k$ eigenvectors have large
amplitude. For the inverse problem of the heat equation there are two
spatial locations that present a good compromise between the
amplitudes of the first and second eigenvectors, namely $x_1$ and
$x_2$ --- see Fig.~\ref{fig:eigenvectors}. We have $m=4$ measurements
at our disposal but only two spatial locations that are a good
compromise between the first and second scaled eigenvectors. Thus,
clusterization arises as a consequence of the pigeonhole principle.

\begin{figure}\label{fig:eigenvectors}
    \centering
    \includegraphics[width=\textwidth]{figs/eigenvectors_dst_scaled.png}
    \caption{D-optimal measurement locations ($m=4$ measurements) and
      weighted eigenvectors for finding the initial condition of the
      1D heat equation. Measurement locations and weighted
      eigenvectors are plotted over the computational domain $\Omega =
      [0, 1]$ (x-axis). Measurement clusterization occurs
      approximately at $0.31$ and $0.69$. These two locations are a
      compromise between the magnitudes of the first and second
      eigenvectors, which are the eigenvectors that a D-optimal design
      aims to measure. Allocating $m=4$ measurements into two
      locations results in clusterization, according to the pigeonhole
      principle.}
  \label{fig:why}
\end{figure}
